---
title: "assignment_1.Rmd"
author: "Bryan Urban"
date: "Thursday, March 12, 2015"
output: html_document
---

This project looks at activity monitoring data. First we will look at the number of steps taken by day.

## Activity Data by Day
```{r}
library(dplyr)
library(lubridate)
df <- read.csv("activity.csv")
df$date <- ymd(df$date)
# summary by date
dfs <- 
  df %>%
  group_by(date) %>%
  summarize(steps=sum(steps, na.rm=TRUE))
hist(dfs$steps)
steps_mean <- mean(dfs$steps)
steps_median <- median(dfs$steps)
```

The mean number of steps is `r steps_mean` and the median number of steps is `r steps_median`.

## Activity Data by Interval
```{r}
dfi <-
  df %>%
  group_by(interval) %>%
  summarize(steps=mean(steps, na.rm=TRUE))
plot(dfi$interval, dfi$steps, type='l')
max_steps_interval <- dfi[which(dfi$steps==max(dfi$steps)),]$interval
```
The interval with the maximum average number of steps is `r max_steps_interval`. 

## Imputing Missing Data 
```{r}
# fill in missing data with approximations from the mean interval steps
df_missing <- df[is.na(df$steps),]
missing_steps <- inner_join(df_missing, dfi, by="interval")$steps.y
df[is.na(df$steps),]$steps <- missing_steps

# recalculate mean values
dfs <- 
  df %>%
  group_by(date) %>%
  summarize(steps=sum(steps, na.rm=TRUE))
hist(dfs$steps)
steps_mean <- mean(dfs$steps)
steps_median <- median(dfs$steps)
```
The new daily mean is `r steps_mean` and median is `r steps_median`. Values have increased slightly. Imputing missing data by interval can bias the mean and median of daily total steps.

## Weekends vs. Weekdays
```{r}
library(lattice)
library(directlabels)
df$weekend <- weekdays(df$date) %in% c("Saturday", "Sunday")
df$weekend <- sapply(df$weekend, function(x) {if(x) "Weekend" else "Weekday"})

dfw <-
  df %>%
  group_by(weekend, interval) %>%
  summarize(steps=mean(steps, na.rm=TRUE))
xyplot(steps~interval|weekend, dfw, type='l')
direct.label(xyplot(steps~interval, dfw, groups=weekend, type='l'))
```
On the weekends, people takemore staps laer in the day - there is a shift in the morning. Perhaps people like to sleep in on the weekends. 


